Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2019-10-07 manuscript/manuscript.tex
changes to median

A long-standing problem in evolutionary genetics is quantifying the roles of
genetic drift and selection in shaping genome-wide allele frequency changes.
Selection can both directly and indirectly affect allele frequencies, with the
indirect effect coming from the action of selection on correlated loci
elsewhere in genome e.g. linked selection (\cite{Maynard_Smith1974-lc},
\cite{Charlesworth1993-gb,Nordborg1996-nq}; see \cite{Barton2000-zg} for a
review). Previous work on this question has mostly focused on teasing apart the
impacts of drift and selection on genome-wide diversity using population
samples from a single contemporary timepoint, often by modeling the correlation
between regional recombination rate, gene density, and diversity created in the
presence of linked selection \parencite{Cutter2013-ba,Sella2009-nx}. This
approach has shown linked selection has a major role in shaping patterns of
genome-wide diversity across the genomes of a range of sexual species
\parencite{Begun2007-bg,Beissinger2016-cm,Sattath2011-dr,Williamson2014-oy,Andersen2012-bj,Cutter2010-gi,Elyashiv2016-vt},
and has allowed us to quantify the relative influence of positive selection
(hitchhiking) and negative selection (background selection;
\cite{Nordborg2005-dc,McVicker2009-ax,Hernandez2011-gs,Elyashiv2016-vt}.
However, we lack an understanding of how genome-wide linked selection acts over
time.

@@

\vb{One concern is that these covariances reflect the localized impact of a few
  large-effect loci rather than selection on a polygenic trait. Since our
  covariances are essentially averages over loci, the covariance estimate can
  be strongly affected by a single outlier region. To test whether large
  outlier regions drive the genome-wide signal we see in the
  \textcite{Barghi2019-qy} data, we calculate the covariances in 100kb windows
  along the genome (we refer to these as windowed covariances throughout). We
  take the median windowed covariance as a measure of the genome-wide
  covariance robust to large-effect loci; we find that the median windowed
  covariance $\cov(\Delta p_0, \Delta p_{10})$ (pooling all replicates) is also
  significantly different from zero, 0.0016 (0.95\% CIS [0.0015, 0.0017], see
  Supplementary Table \ref{supp:table-trimmed-mean} for more timepoints and
  trimmed means). This median covariance is also limiting result of a trimmed
  mean that excludes the upper and lower $\alpha$ tails to calculate the
  trimmed average windowed covariance. As $\alpha$ increases to 0.5, the
  trimmed covariance converges to the median windowed covariance (by the
  definition of the median; see Supplementary Figure
  \ref{suppfig:barghi-trimmed-mean}). Thus our genomic temporal covariances are
  non-zero due to the impact of selection on many genomic windows.  }

>>>>

A long-standing problem in evolutionary genetics is quantifying the roles of
genetic drift and selection in shaping genome-wide allele frequency changes.
Selection can both directly and indirectly affect allele frequencies, with the
indirect effect coming from the action of selection on correlated loci
elsewhere in genome e.g. linked selection (\cite{Maynard_Smith1974-lc},
\cite{Charlesworth1993-gb,Nordborg1996-nq}; see \cite{Barton2000-zg} for a
review). Previous work on this question has mostly focused on teasing apart the
impacts of drift and selection on genome-wide diversity using population
samples from a single contemporary timepoint, often by modeling the correlation
between regional recombination rate, gene density, and diversity created in the
presence of linked selection \parencite{Cutter2013-ba,Sella2009-nx}. This
approach has shown linked selection has a major role in shaping patterns of
genome-wide diversity across the genomes of a range of sexual species
\parencite{Begun2007-bg,Beissinger2016-cm,Sattath2011-dr,Williamson2014-oy,Andersen2012-bj,Cutter2010-gi,Elyashiv2016-vt},
and has allowed us to quantify the relative influence of positive selection
(hitchhiking) and negative selection (background selection;
\cite{Nordborg2005-dc,McVicker2009-ax,ANDOLFATO,MACPHERSON,Hernandez2011-gs,Elyashiv2016-vt}.
However, we lack an understanding of how genome-wide linked selection acts over
time.

@@

\vb{One concern is that these covariances reflect the localized impact of a few
  large-effect loci rather than selection on a polygenic trait. Since our
  covariances are essentially averages over loci, the covariance estimate can
  be strongly affected by a single outlier region. To test whether large
  outlier regions drive the genome-wide signal we see in the
  \textcite{Barghi2019-qy} data, we calculate the covariances in 100kb windows
  along the genome (we refer to these as windowed covariances
  throughout) and take the median windowed covariance, and
  trimmed-mean windowed covariance, as a measure of the genome-wide
  covariance robust to large-effect loci. These robust estimates (Supplementary Table \ref{supp:table-trimmed-mean})
  confirm the patterns we see using the mean covariance, confirming that genomic temporal covariances are
 non-zero due to the impact of selection on many genomic windows.

%  This median covariance is also limiting result of a trimmed
 % mean that excludes the upper and lower $\alpha$ tails to calculate the
 % trimmed average windowed covariance. As $\alpha$ increases to 0.5, the
 % trimmed covariance converges to the median windowed covariance (by the
 % definition of the median; see Supplementary Figure
 % \ref{suppfig:barghi-trimmed-mean}). Thus our genomic temporal covariances are
 % non-zero due to the impact of selection on many genomic windows.  }

%We
%  take the median windowed covariance as a measure of the genome-wide
 % covariance robust to large-effect loci; we find that the median windowed
 % covariance $\cov(\Delta p_0, \Delta p_{10})$ (pooling all replicates) is also
 % significantly different from zero, 0.0016 (0.95\% CIS [0.0015, 0.0017], see
 % Supplementary Table \ref{supp:table-trimmed-mean} for more timepoints and
 % trimmed means).
